Discovering governing reactions from concentration data

### 2018-01-25 docs/manuscript.tex
f

\usepackage[left=2.5cm,right=2.6cm,top=3cm,bottom=3cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[intlimits]{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{subcaption}
\usepackage[version=4]{mhchem}

@@

@@

	\section{Introduction}
	When presented with a time series of possibly noisy non-equilibrium concentration fluctuations of some species as output of, e.g., measurements from experiments or simulations that were parameterized by microscopic rates \textcolor{red}{(cite ReaDDy?)}, one can ask for the corresponding macroscopic rates and a generating reaction network.
	In this paper we present an application of the shallow learning method SINDy \cite{Brunton2015}. By sparse regression, it is able to identify generating nonlinear dynamics in data that stems from dynamical systems. The parsimonious nature of the results avoids overfitting and provides interpretability.
	In our application we, as opposed to the original method, do not only look for macroscopic rates of net species change but investigate the specific reactions that might have lead to the observations.
	We demonstrate the algorithm on two toy problems - one problem showing that when there is no ambiguity in the system, one converges to the correct rates with increasing resolution of concentration fluctuations and one in which we compute a sparse reaction network for given data.

@@

	When presented with multiple trajectories one can extend the objective function by a sum over these and obtain a more general minimization problem that reads: Find $\hat{\Xi}$ such that
	\begin{align}
		\hat{\Xi} = \underset{\Xi}{\arg\min}\left( \sum_{\text{traj}}\frac{1}{2T_{\text{traj}}}\left\| \dot{\textbf{X}}_\text{traj} - \Theta(\textbf{X}_\text{traj})\Xi \right\|_F^2 + \alpha\lambda\|\Xi\|_1 + \alpha(1-\lambda)\|\Xi\|_2^2 \right)
	\end{align}
	subject to $\Xi\geq 0$.
	For $\alpha=0$ the minimization problem reduces to constrained least-squares. In order to solve (\ref{method:minimizationproblem}) we apply the sequential least-squares minimizer SLSQP, originally described in \cite{Kraft1988}, by the software package SciPy \cite{SciPy}. As initial guess for the regularized problem we use the solution of (\ref{method:minimizationproblem}) with $\alpha=0$.

@@

    To demonstrate our method we estimate a gene-regulatory network from data. The data is time-series of molecule-concentrations. We generate the data by integrating the law of mass action (LMA) equations of the true model and add gaussian noise.

    We describe the gene-regulatory network as follows. Let A, B and C be the three species of proteins. Each of them is translated from its according mRNA molecule. Each mRNA molecule is transcribed from its according DNA. The proteins and mRNA molecules decay over time. For each species one can formulate the following basic reactions
    \begin{align*}
        \ce{ DNA_i } & \ce{ -> DNA_i + mRNA_i } &\text{(transcription)} \\
        \ce{ mRNA_i } & \ce{ -> mRNA_i + Protein_i } &\text{(translation)} \\
        \ce{ mRNA_i } & \ce{ -> } \emptyset &\text{(decay of mRNA)} \\
        \ce{ Protein_i } & \ce{ -> } \emptyset &\text{(decay of Protein)}
    \end{align*}

    A, B and C regulate each other in a negative way, by hindering the transcription process. In the LMA model we account for this by a reaction. A repression of species $j$ by species $i$ is modeled as
    \begin{equation*}
        \ce{ Protein_i + mRNA_j -> Protein_i } \quad \text{(regulation by repression)}
    \end{equation*}

    In our example A proteins regulate the mRNA$_\mathrm{B}$ molecules, B proteins regulate the mRNA$_\mathrm{A}$ molecules and C proteins regulate the mRNA$_\mathrm{A}$ molecules. The network of all species and reactions is depicted in figure \ref{fig:network}. This will serve as our reference model to generate the time-series of concentrations.
    \begin{figure}
        \centering
        \includegraphics[width=.5\textwidth]{./figures_tex/gene_regulation_full.pdf}
        \caption{The regulation network example described in Section \ref{sec:generegulation}. Each circle is a species, each arrow corresponds to one reaction. The red arrows describe the regulatory network which generates our data.}
        \label{fig:network}
    \end{figure}

    \section{Scenarios of measurement}

    The proposed estimation method is designed to find reaction networks and rates from experimental data. Therefore one can define the following measurement scenarios, that differentiate how many timeseries are available and how much noise is contained in the data:
    \begin{enumerate}
        \item The initial state of the measurement apparatus can be reproduced exactly. There is only one set of generating reaction rates. Each datapoint is subject to noise. The outcome of this scenarios is one time-series of concentrations with a certain noise-level.
        \item The process is observed multiple times. The initial state of the measurement apparatus differs in each observation. The outcome of this scenario is multiple time-series each with a given noise level and different initial states.
        \item The process is observed multiple times. The initial state of the measurement apparatus differs in each observation. Additionally the generating reaction rates slightly differ in each observation. The outcome of this scenario is multiple time-series each with a given noise level, different initial states and slighty perturbed underlying reaction rates.
    \end{enumerate}

@@

	For generating time series data of concentrations we integrate the law of mass action equations and add Gaussian noise. In that manner we produce several realizations for a couple of initial conditions. The derivatives are evaluated by finite differences of first order.

@@

@@

	The problem now is to find a suitable $\alpha$ such that a parsimonious solution is yielded. To this end, cross-validation is applied. The trajectories that are used as training and test data set are depicted in Figure \ref{ex2:test-train-concentration-curves}. \textcolor{red}{Perhaps use several test trajectories with different initial conditions to get something like a confidence interval?} As an indicator to the estimation error one can use the scoring function
	\begin{align}
		f_\alpha = \frac{1}{T}\|\mathbf{\dot{X}}_\text{test} - \Theta (\textbf{X}_\text{test}) \Xi_\text{train}\|_F^2,
		\label{ex2:score-function}
	\end{align}
	where the best possible score is $f_\alpha = 0$ and larger values indicate worse models.
	For a selection of hyperparameters $\alpha\in [0, 100]$, the model parameters $\Xi$ are estimated and the scoring function (\ref{ex2:score-function}) evaluated, yielding a dependency of the score on $\alpha$ as depicted in Figure \ref{ex2:score-graph}.

>>>>

\usepackage[left=2.5cm,right=2.6cm,top=3cm,bottom=3cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[intlimits]{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage[version=4]{mhchem}

@@

\newsavebox{\selvestebox}
\newenvironment{colbox}[1]
{\newcommand\colboxcolor{#1}%
	\begin{lrbox}{\selvestebox}%
		\begin{minipage}{\dimexpr\columnwidth-2\fboxsep\relax}}
		{\end{minipage}\end{lrbox}%
	\begin{center}
		\colorbox[HTML]{\colboxcolor}{\usebox{\selvestebox}}
\end{center}}
\newcommand\restr[2]{{% we make the whole thing an ordinary symbol
		\left.\kern-\nulldelimiterspace % automatically resize the bar with \right
		#1 % the function
		\vphantom{\big|} % pretend it's a little taller at normal size
		\right|_{#2} % this is the delimiter
}}

@@

	\section{Introduction}
	When presented with a time series of possibly noisy non-equilibrium concentration fluctuations of some species as output of, e.g., measurements from experiments or simulations that were parameterized by microscopic rates, one can ask for the corresponding macroscopic rates and a generating reaction network.
	In this paper we present an application of the shallow learning method SINDy \cite{Brunton2015}. By sparse regression, it is able to identify generating nonlinear dynamics in data that stems from dynamical systems. The parsimonious nature of the results avoids overfitting and provides interpretability.
	In our application we, as opposed to the original method, do not only look for macroscopic rates of net species change but investigate the specific reactions that might have lead to the observations.
	We demonstrate the algorithm on two toy problems - one problem showing that when there is no ambiguity in the system, one converges to the correct rates with increasing resolution of concentration fluctuations and one in which we compute a sparse reaction network for given data.

@@

	When presented with multiple trajectories one can extend the objective function by a sum over these and obtain a more general minimization problem that reads: Find $\hat{\Xi}$ such that
	\begin{align}
		\hat{\Xi} = \underset{\Xi}{\arg\min}\left( \sum_{\text{traj}}\frac{1}{2T_{\text{traj}}}\left\| \dot{\textbf{X}}_\text{traj} - \Theta(\textbf{X}_\text{traj})\Xi \right\|_F^2 + \alpha\lambda\|\Xi\|_1 + \alpha(1-\lambda)\|\Xi\|_2^2 \right)
	\end{align}
	subject to $\Xi\geq 0$.
	For $\alpha=0$ the minimization problem reduces to constrained least-squares. In order to solve (\ref{method:minimizationproblem}) the sequential least-squares minimizer SLSQP \cite{Kraft1988} is applied via the software package SciPy \cite{SciPy}. Since only the concentration data $\mathbf{X}$ is available but not its temporal derivative $\dot{\mathbf{X}}$, it is approximated numerically by second order finite differences with the exception of boundary data.

@@

    To demonstrate the method a gene-regulatory network is estimated from time series of molecule-concentrations. The data of is generated by integrating the law of mass action equations of the underlying model and adding Gaussian noise.

    To this end, let $A$, $B$, and $C$ be three species of proteins which are being translated from a corresponding mRNA molecule. Each mRNA in turn has a corresponding DNA which it is transcribed from. The proteins and mRNA molecules decay over time whereas the DNA concentration remains constant.

    For each of the three species one can formulate the following basic reactions
	\begin{align*}
        \ce{ DNA_i } & \ce{ -> DNA_i + mRNA_i } &\text{(transcription)}, \\
        \ce{ mRNA_i } & \ce{ -> mRNA_i + Protein_i } &\text{(translation)}, \\
        \ce{ mRNA_i } & \ce{ -> } \emptyset &\text{(decay of mRNA)}, \\
        \ce{ Protein_i } & \ce{ -> } \emptyset &\text{(decay of Protein)},
    \end{align*}
    for $i=A,B,C$.

    The protein concentrations of $A$, $B$, and $C$ regulate each other in a negative way by hindering the transcription process. In the law of mass action model we account for this by a reaction. A repression of species $j$ by species $i$ can be modeled as
    \begin{equation*}
        \ce{ Protein_i + mRNA_j -> Protein_i } \quad \text{(regulation by repression)}.
    \end{equation*}

    In our example proteins of type $A$ regulate the mRNA$_\mathrm{B}$ molecules, proteins of type $B$ regulate the mRNA$_\mathrm{A}$ molecules and proteins of type $C$ regulate the mRNA$_\mathrm{A}$ molecules. The network of all species and reactions is depicted in Figure \ref{fig:network}. This serves as reference model to generate the time series of concentrations.
    \begin{figure}
        \centering
        \includegraphics[width=.5\textwidth]{./figures_tex/gene_regulation_full.pdf}
        \caption{The regulation network example described in Section \ref{sec:generegulation}. Each circle depicts a species, each arrow corresponds to one reaction. The red arrows describe the regulatory network which generates the example data.}
        \label{fig:network}
    \end{figure}

    \section{Scenarios of measurement}\label{sec:scenarios-of-measurement}

    The proposed estimation method is designed to find reaction networks and rates from experimental data. In the following measurement scenarios are considered that differ in how many time series are available and the level of noise that the data is subject to:
    \begin{enumerate}
        \item The initial state of the measurement apparatus can be reproduced exactly. There is only one set of generating reaction rates. Each frame of a time series is subject to noise. The outcome of this scenario is one time series of concentrations with a certain noise-level.
        \item The process is observed multiple times. The initial state of the measurement apparatus differs in each observation. The outcome of this scenario is multiple time series each with a certain noise level and different initial states.
        \item The process is observed multiple times. The initial state of the measurement apparatus differs in each observation. Additionally the generating reaction rates differ slightly in each observation. The outcome of this scenario is multiple time series each with a given noise level, different initial states and slightly perturbed underlying reaction rates.
    \end{enumerate}

@@

	In the following two examples are presented. The first example considers data according to the first scenario of measurement as presented in Section \ref{sec:scenarios-of-measurement}.
	The minimization problem (\ref{method:minimizationproblem}) with $\alpha=0$ is solved for different magnitudes of noise variance.

@@

	In this example it is demonstrated that given enough data, the original model can be recovered in the limit of the noise variance $\sigma^2\to 0$.

	\begin{colbox}{F8E0E0}
		todo
	\end{colbox}

@@

	The problem now is to find a suitable $\alpha$ such that a parsimonious solution is yielded. To this end, cross-validation is applied. The trajectories that are used as training and test data set are depicted in Figure \ref{ex2:test-train-concentration-curves}. \textcolor{red}{Perhaps use several test trajectories with different initial conditions to get something like a confidence interval?} As an indicator to the estimation error one can use the scoring function
	\begin{align}
		f_\alpha = \frac{1}{T}\|\dot{\mathbf{X}}_\text{test} - \Theta (\textbf{X}_\text{test}) \Xi_\text{train}\|_F^2,
		\label{ex2:score-function}
	\end{align}
	where the best possible score is $f_\alpha = 0$ and larger values indicate worse models.
	For a selection of hyperparameters $\alpha\in [0, 100]$, the model parameters $\Xi$ are estimated and the scoring function (\ref{ex2:score-function}) evaluated, yielding a dependency of the score on $\alpha$ as depicted in Figure \ref{ex2:score-graph}.
